Dust trapping in thin-ringed protoplanetary disks

### 2018-06-25 projects/dullemond/paper_dusttraps/ms.tex
New version of paper on dust traps by Kees et al.

%\documentclass[referee]{aa}
\documentclass{aa}
%\documentclass{mn2e}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\def\remark#1{{{\bf remark:} \bf #1}}
\def\action#1{{\bf #1}}
\def\putin#1{{\it #1}}
\def\revised#1{{\bf #1}}
\newcommand{\todo}[1]{\textcolor{red}{\bf #1}}
%
\def\aap{A\&A}
\def\apss{A\&SS}
\def\pasp{Proc.Astr.Soc.Pacific}
\def\pasj{Publ.Astr.Soc.Japan}
\def\araa{AnnRevA\& A}
\def\mnras{MNRAS}
\def\icarus{Icarus}
\def\nat{Nature}
\def\apj{ApJ}
\def\aj{AJ}
\def\apjl{ApJL}
\def\apjs{ApJS}
\def\jqsrt{JQSRT}
\def\max{\mathrm{max}}
\def\erg{\hbox{erg}}
\def\sec{\hbox{s}}
\def\Hz{\hbox{Hz}}
\def\gram{\hbox{g}}
\def\cm{\hbox{cm}}
\def\km{\hbox{km}}
\def\kel{\hbox{K}}
\def\KeV{\hbox{KeV}}
\def\MeV{\hbox{MeV}}
\def\eV{\hbox{eV}}
\def\yr{\hbox{yr}}
\def\AU{\hbox{AU}}
\def\ster{\hbox{ster}}
%\def\in{\mathrm{in}}
\def\out{\mathrm{out}}
\def\eff{\mathrm{eff}}
\def\rad{\mathrm{rad}}
\def\vrt{\mathrm{vert}}
\def\vset{\ensuremath{v_\mathrm{sett}}\xspace}
\def\fact{\ensuremath{\xi}\xspace}
\def\cs{\ensuremath{c_\mathrm{s}}\xspace}
\def\siggas{\ensuremath{\sigma_\mathrm{g}}\xspace}
\def\sigcoag{\ensuremath{\sigma_\mathrm{c}}\xspace}
\def\Tgas{\ensuremath{T_\mathrm{gas}}\xspace}
%\def\rhogas{\ensuremath{\rho_\mathrm{gas}}\xspace}
\def\rhogas{\ensuremath{\rho}\xspace}
\def\rhodust{\ensuremath{\rho_\mathrm{d}}\xspace}
\def\mugas{\ensuremath{\mu_\mathrm{gas}}\xspace}
\def\mp{\ensuremath{m_\mathrm{p}}\xspace}
\def\Hp{\ensuremath{H_\mathrm{p}}\xspace}
\def\Hs{\ensuremath{H_\mathrm{s}}\xspace}
\def\Omegak{\ensuremath{\Omega_\mathrm{K}}\xspace}
\def\Mstar{\ensuremath{M_\star}\xspace}
\def\Sc{\ensuremath{\mathrm{Sc}}\xspace}
\def\St{\ensuremath{\mathrm{St}}\xspace}
\def\Pt{\ensuremath{\mathrm{Pt}}\xspace}
\def\comma{\,,}
\def\fullstop{\,.}
%
\def\submitted{submitted\xspace}
\def\inprep{in prep.\xspace}
%
\begin{document}
%\thesaurus{02.01.2,08.03.4,08.06.2,08.16.5,13.09.6}
\title{Dust trapping in thin-ringed protoplanetary disks}  \titlerunning{Thin-ringed disks}
\authorrunning{Dullemond, Birnstiel, Perez, Andrews, Isella, Huang, Zhu, Guzman, Benisty, Carpenter}
\author{C.P.~Dullemond$^{1}$, T.~Birnstiel$^{2}$, L.~Perez$^{3}$, S.~Andrews$^{4}$, A.~Isella$^{5}$, J.~Huang$^{4}$, Z.~Zhu$^{6}$, V.~Guzm\'an$^{7}$, M.~Benisty$^{8}$, J.~Carpenter$^{9}$}
\institute{
  (1) Zentrum f\"ur Astronomie, Heidelberg University, Albert Ueberle Str.~2, 69120 Heidelberg, Germany\\
  (2) {\bf XXXX}
  (3) {\bf XXXX}
  (4) {\bf XXXX}
  (5) {\bf XXXX}
  (6) {\bf XXXX}
} \date{\today}

@@

\section{Introduction}
The concept of dust trapping in local pressure maxima has become a central theme
in studies of planet formation and protoplanetary disk evolution, as it may
provide an elegant solution to several problems in these fields of
study. Theories of planet formation are plagued by the ``radial drift barrier'':
the problem that as dust aggregates grow by coagulation, they tend to radially
drift toward the star before they reach planetsimal size
\citep[e.g.][]{2010A&A...513A..79B}. A natural solution to this problem could be
the trapping of dust particles in local pressure maxima
\citep{1972fpp..conf..211W, 2007ApJ...664L..55K, 1995A&A...295L...1B,
  1997Icar..128..213K}. From an observational perspective, the radial drift
problem manifests itself by the presence of large grains in the outer regions of
protoplanetary disks \citep{2003A&A...403..323T, 2009ApJ...700.1502A}, which
appears to be in conflict with theoretical predictions
\citep{2007A&A...469.1169B}. One possible solution to this observational
conundrum could be that the disks are much more massive in the gas than
previously suspected, leading to a higher gas friction for millimeter grains and
thus longer drift times \citep{2017ApJ...840...93P}. But another explanation is
to invoke, again, pressure traps. The most striking observational evidence for
dust trapping seems to come from large transitional disks, which feature giant
dust rings, sometimes lopsided, in which large quantities of dust appears to be
concentrated \citep{2013Natur.493..191C, 2013Sci...340.1199V}. These observations
appear to be well explained by the dust trapping scenario
\citet{2012A&A...545A..81P}. But these transitional disks seem to be rather
violent environments, possibly with strong warps \citep{2015ApJ...798L..44M,
  2017A&A...597A..42B} and companion-induced spirals
\citep{2016ApJ...816L..12D}. For more ``normal'' protoplanetary disk the dust
traps would have to be more subtle. \cite{2012A&A...538A.114P} explored the
possibility that the disk features many axisymmetric local pressure maxima, and
calculated how the dust drift and growth would behave under such conditions. It
was found that, if the pressure bumps are strong enough, the dust trapping can
keep a sufficient fraction of the dust mass at large distances from the star to
explain the observed dust millimeter flux. It would leave, however, a detectable
pattern of rings that should be discernable with ALMA observations.  Since the
multi-ringed disk observation of HL Tau \citep{2015ApJ...808L...3A} a number of
such multi-ringed disks have been detected \citep{2016ApJ...820L..40A,
  2016PhRvL.117y1101I, 2017ApJ...851L..23C, 2017A&A...600A..72F,
  2018A&A...610A..24F}. It is therefore very tempting to see also these
multi-ringed disks as evidence for dust trapping, and as an explanation for
the retention of dust in the outer regions of protoplanetary disks.

@@

If we assume that the rings seen in our data are caused by dust trapping, the
question arises: what constraints do the data impose on the physics of dust
trapping? To explore this we start by studying the rings individually, assuming
that the dust does not escape from the ring. This makes it possible to look for
a steady-state dust trapping model in which the radial drift forces (that push
the dust to the pressure peak) are balanced by turbulent mixing (that tends to
smear out the dust away from the pressure peak). In Section
\ref{sec-steady-state-analytic-trap-model} we will construct a very simplified
analytic dust trapping model, and confront this with the most well-isolated
rings from our sample. {\bf Describe the next sections.} But before we do the
modeling, we review, in Section \ref{sec-data}, the key aspects of the data from
our ALMA Large Programme that we compare our models to.

\section{The high-contrast rings of AS 209, Elias 24, GW Lup, HD 143006 and HD 163296}
\label{sec-data}
%
In this paper we discuss in particular those sources of the ALMA Large Programme
that show high-contrast and radially thin rings that are separated by deep
valleys, and that are sufficiently face-on to not have to worry about 3-D
lie-of-sight aspects. These are AS 209, Elias 24, GW Lup, HD 143006 and HD
163296. The high-contrast rings of these sources provide ``clean laboratories''
for testing the theory of dust trapping, both in a ring-by-ring manner, as well
as by comparing to global dust drift models. The most striking case is AS 209,
which features two high-contrast radially thin rings which are isolated and
large ({\bf see Guzm\'an et al.~(2018) for the detailed description of this
  dataset}). Fig.\ref{fig-obs-as209} shows the radial profile (deprojected for
inclination) of the thermal emission of the dust of the disk around AS 209. The
brightness is expressed as linear brightness temperature
($T_{\mathrm{bright,linear}}=(c^2/2k_B)I_\nu/\nu^2$), where $I_\nu$ is the
intensity, $c$ the light speed, $k_B$ the Boltzmann constant, and $\nu$ the
frequency. It should be noted that the full (non-linear) brightness temperature
(defined such that $B_\nu(T_{\mathrm{bright,full}})=I_\nu$, with $B_\nu(T)$ the
Planck function) can have very different values, in particular at low
brightness. As a result $T_{\mathrm{bright,full}}(r)$ tends to vastly
underestimate the contrast between the rings and the inter-ring
regions. Therefore, in this paper we plot the brightness using the linear
brightness temperature.

\begin{figure}
\centerline{\includegraphics[width=0.5\textwidth]{as209_gauss_modelfit.pdf}}
\caption{\label{fig-obs-as209}The linear brightness temperature in band 6 of the
  disk around AS 209, as a function of deprojected distance to the star (see
  {\bf Guzm\'an et al.~(2018)} for a detailed description and analysis of these data). The
  vertical axis is logarithmic to better show the contrast. The two most
  prominent rings are fitted by Gaussian rings. Ring 1 is fitted by
  $T_{\mathrm{bright,linear}}(r)=
  3.5\,\exp(-0.5(r-77\,\mathrm{au})^2/(4.0\,\mathrm{au})^2)$, while ring 2 is fitted by
  $T_{\mathrm{bright,linear}}(r)=
  2.8\,\exp(-0.5(r-124.7\,\mathrm{au})^2/(4.4\,\mathrm{au})^2)$.}
\end{figure}

The two most prominent rings (seen as bumps in Fig.\ref{fig-obs-as209}) are
fitted by Gaussian profiles with a standard deviation width of
$4.0\,\mathrm{au}$ and $4.4\,\mathrm{au}$ respectively. This corresponds to a
full-width-at-half-maximum (FWHM) of $9.3\,\mathrm{au}$ and $10.3\,\mathrm{au}$.
The ALMA beam FWHM for these observations is roughly $4.4\,\mathrm{au}$, meaning
that the rings are 2.1 and 2.3 beams wide.

These Gaussian profiles fit the observed data remarkably well, even in the wings
of the profile (except the outer wing of the outer ring). This is somewhat
surprising, given that the rings are resolved. The Gauss profiles therefore do
not represent the beam profile, but tell something about the actual dust
distribution within the ring.

The source Elias 24 also shows a prominent high-contrast isolated ring,
albeit flanked on the outside by a plateau of emission (see {\bf Andrews
  et al.~(2018) for a detailed description of this dataset}).

\begin{figure}
\centerline{\includegraphics[width=0.5\textwidth]{elias24_gauss_modelfit.pdf}}
\caption{\label{fig-obs-elias24}As Fig.~\ref{fig-obs-as209}, but now for the
  source Elias 24 (see {\bf Andrews et al.~(2018)} for a detailed description
  and analysis of these data). The most prominent ring is fitted by a Gaussian
  profile $T_{\mathrm{bright,linear}}(r)=
  5.3\,\exp(-0.5(r-79.9\,\mathrm{au})^2/(5.0\,\mathrm{au})^2)$.}
\end{figure}

The ring can be reasonably well fitted with a Gaussian profile with standard
deviation width of $5.0\,\mathrm{au}$ corresponding to a FWHM of
$11.7\,\mathrm{au}$. Given a beam size in band 6 of about $4.9\,\mathrm{au}$,
the ring of Elias 24 is also resolved by about 2.4 beam widths. The Gaussian fit
is, however, less convincing in the case of Elias 24 than in the case of AS 209,
so that it cannot be clearly argued that the ring has a Gaussian shape in the
case of Elias 24.

{\bf [TODO: GW Lup, HD 143006 and HD 163296]}

\section{Steady-state dust distribution in a ringlike trap}
\label{sec-steady-state-analytic-trap-model}
%
Given the nicely separated narrow rings in the above sources, we can analyze
these rings on a one-by-one basis. This is what we will do in this Section. We
will assume that the dust trapping in these rings is absolute: no dust escapes.
But increased turbulent mixing can spread the dust over a broader ring. Before
we start constructing a model, we will discuss an important feature of the
observed rings: that they appear to be radially resolved, Gaussian in shape, yet
sub-Planckian in brightness. We will present a simple analytic dust trapping
model, and show which radial brightness profiles it predicts, depending on the
dust size distribution, the turbulent strength and the gas pressure bump width.
This, when compared to the data, will allow us to draw some conclusions about
the size distribution {\bf [and perhaps more?]}.

\subsection{Resolved ring width and the brightness temperature}
If our finding is correct that the rings in our sample are all spatially
resolved in radial direction, then this poses an interesting question. In many
cases the brightness of these rings is arguably sub-Planckian. This is, for
instance, the case for the high-contrast rings of AS 209, as well as {\bf
  XXXXXXX}. For the outer two rings of AS 209 (at 77 au and 125 au,
respectively, the non-linear brightness temperatures are $3.5\,\mathrm{K}$ and
$2.8\,\mathrm{K}$, respectively. If these rings would have been {\em
  un}resolved, then this sub-Planckian emission is easily explained by the ring
not filling the beam. But because these rings are radially resolved, this
scenario should be rejected. The remaining possibilities are that the disk is
colder than we think or that the rings are optically thin. In the latter case
the spectral slope $\alpha_{\mathrm{spec}}$ at the location of these rings, once
measured, will determine if the grains that are trapped in this ring are
smaller, similar or larger than about $\lambda/2\pi$, where $\lambda$ is the
wavelength around which the slope is measured. If we use
$\lambda=0.13\,\mathrm{cm}$ and $\lambda=0.31\,\mathrm{cm}$ (representing ALMA
band 6 and 3, respectively) to determine $\alpha_{\mathrm{spec}}$ this would
mean we can distinguish between grains smaller, similar or larger than about 300
$\mu$m. A low $\alpha_{\mathrm{spec}}$ would imply grains that are substantially
larger than 300 $\mu$m, i.e.~of the order of millimeter or larger.

In several recent observations of the spectral index across ringed disks
\citep{2015ApJ...808L...3A,2018ApJ...852..122H} one clearly sees that
$\alpha_{\mathrm{spec}}$ varies across these rings, being closer to $2$ at the
ring center and substantially larger between the rings. This makes sense in
terms of the dust trapping scenario in which we expect larger grains to be
trapped more efficiently (and thus dominate the peak of the ring) than smaller
grains, because the smaller grains will be more subject to turbulent
mixing. This scenario requires a grain size distribution, so that the width of
the dust ring is smaller for the bigger grains and bigger for the smaller ones.

While this scenario is plausible and appealing, one may worry that it requires
some fine-tuning. Why would the rings be dominated by a population of grains
that are larger than 300 $\mu$m, but not too large, for else they would be so
effectively trapped that they would produce an even narrower ring? For one or
two sources that could be coincidence, but for a multitude of sources this is
less plausible. One possible explanation is that millimeter grains are indeed
the largest grains in the outer regions protoplanetary disks, because larger
grains would easily get destroyed and/or simply bounce instead of stick when
they collide, and thus not grow further \citep{2010A&A...513A..56G,
  2010A&A...513A..57Z}. These models also predict larger grains to drift quickly
inward, but that is only valid for a model without dust traps. At any rate:
perhaps millimeter grains are a natural consequence of dust growth. If so, it
then remains to be understood why the combination of turbulent strength and gas
surface density conspire to trap these millimeter grains, but not too much, for
else we might expect at least some rings to be narrower than a pressure scale
height. We are aware that, even with a sample of 20 sources, we are still in the
small-number-statistics regime, so these considerations are limited in their
strength.

In this section we will study this question using a very simple model of a dust
trap: a parameterized gaussian pressure profile. The goal is to find out if the
variations of $\alpha_{\mathrm{spec}}$ across the ring are naturally produced,
while still keeping the dust ring radially resolved, without having to fine-tune
the size distribution of the grains, the turbulence of the gas or other
parameters. We will do this in several steps, starting with analytic solutions
in Subsection \ref{sec-analytic-trap}, and proceeding with DISKLAB models in
Subsection \ref{sec-disklab-trap}.

>>>>

%\documentclass[referee]{aa}
\documentclass{aa}
%\documentclass{mn2e}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\def\remark#1{{{\bf remark:} \bf #1}}
\def\action#1{{\bf #1}}
\def\putin#1{{\it #1}}
\def\revised#1{{\bf #1}}
\newcommand{\todo}[1]{\textcolor{red}{\bf #1}}
%
\def\aap{A\&A}
\def\apss{A\&SS}
\def\pasp{Proc.Astr.Soc.Pacific}
\def\pasj{Publ.Astr.Soc.Japan}
\def\araa{AnnRevA\& A}
\def\mnras{MNRAS}
\def\icarus{Icarus}
\def\nat{Nature}
\def\apj{ApJ}
\def\aj{AJ}
\def\apjl{ApJL}
\def\apjs{ApJS}
\def\jqsrt{JQSRT}
\def\max{\mathrm{max}}
\def\erg{\hbox{erg}}
\def\sec{\hbox{s}}
\def\Hz{\hbox{Hz}}
\def\gram{\hbox{g}}
\def\cm{\hbox{cm}}
\def\km{\hbox{km}}
\def\kel{\hbox{K}}
\def\KeV{\hbox{KeV}}
\def\MeV{\hbox{MeV}}
\def\eV{\hbox{eV}}
\def\yr{\hbox{yr}}
\def\AU{\hbox{AU}}
\def\ster{\hbox{ster}}
%\def\in{\mathrm{in}}
\def\out{\mathrm{out}}
\def\eff{\mathrm{eff}}
\def\rad{\mathrm{rad}}
\def\vrt{\mathrm{vert}}
\def\vset{\ensuremath{v_\mathrm{sett}}\xspace}
\def\fact{\ensuremath{\xi}\xspace}
\def\cs{\ensuremath{c_\mathrm{s}}\xspace}
\def\siggas{\ensuremath{\sigma_\mathrm{g}}\xspace}
\def\sigcoag{\ensuremath{\sigma_\mathrm{c}}\xspace}
\def\Tgas{\ensuremath{T_\mathrm{gas}}\xspace}
%\def\rhogas{\ensuremath{\rho_\mathrm{gas}}\xspace}
\def\rhogas{\ensuremath{\rho}\xspace}
\def\rhodust{\ensuremath{\rho_\mathrm{d}}\xspace}
\def\mugas{\ensuremath{\mu_\mathrm{gas}}\xspace}
\def\mp{\ensuremath{m_\mathrm{p}}\xspace}
\def\Hp{\ensuremath{H_\mathrm{p}}\xspace}
\def\Hs{\ensuremath{H_\mathrm{s}}\xspace}
\def\Omegak{\ensuremath{\Omega_\mathrm{K}}\xspace}
\def\Mstar{\ensuremath{M_\star}\xspace}
\def\Sc{\ensuremath{\mathrm{Sc}}\xspace}
\def\St{\ensuremath{\mathrm{St}}\xspace}
\def\Pt{\ensuremath{\mathrm{Pt}}\xspace}
\def\comma{\,,}
\def\fullstop{\,.}
%
\def\submitted{submitted\xspace}
\def\inprep{in prep.\xspace}
%
\begin{document}
%\thesaurus{02.01.2,08.03.4,08.06.2,08.16.5,13.09.6}
\title{Dust trapping in thin-ringed protoplanetary disks}  \titlerunning{Thin-ringed disks}
\authorrunning{Dullemond, Birnstiel, Troncoso, Perez, Andrews, Isella, Huang, Zhu, Guzman, Benisty, Carpenter}
\author{C.P.~Dullemond$^{1}$, T.~Birnstiel$^{2}$, N.~Troncoso$^{3}$, L.~Perez$^{3}$, S.~Andrews$^{4}$, A.~Isella$^{5}$, J.~Huang$^{4}$, Z.~Zhu$^{6}$, V.~Guzm\'an$^{7}$, M.~Benisty$^{8}$, J.~Carpenter$^{9}$}
\institute{
  (1) Zentrum f\"ur Astronomie, Heidelberg University, Albert Ueberle Str.~2, 69120 Heidelberg, Germany\\
  (2) {\bf XXXX}
  (3) {\bf XXXX}
  (4) {\bf XXXX}
  (5) {\bf XXXX}
  (6) {\bf XXXX}
} \date{\today}

@@

\section{Introduction}
The concept of dust trapping in local pressure maxima has become a central theme
in studies of planet formation and protoplanetary disk evolution, because it might
provide an elegant solution to several problems in these fields of
study. Theories of planet formation are plagued by the ``radial drift barrier'':
the problem that, as dust aggregates grow by coagulation, they tend to radially
drift toward the star before they reach planetsimal size
\citep[e.g.][]{2010A&A...513A..79B}. A natural solution to this problem could be
the trapping of dust particles in local pressure maxima
\citep{1972fpp..conf..211W, 2007ApJ...664L..55K, 1995A&A...295L...1B,
  1997Icar..128..213K}. Not only does this process prevent excessive radial
drift of dust particles, it also tends to concentrate the dust into small
volumes and high dust-to-gas ratios, which is beneficial to planet formation.
From an observational perspective, the radial drift
problem manifests itself by the presence of large grains in the outer regions of
protoplanetary disks \citep{2003A&A...403..323T, 2009ApJ...700.1502A}, which
appears to be in conflict with theoretical predictions
\citep{2007A&A...469.1169B}. One possible solution to this observational
conundrum could be that the disks are much more massive in the gas than
previously suspected, leading to a higher gas friction for millimeter grains and
thus longer drift time scales \citep{2017ApJ...840...93P}. Another explanation is
to invoke dust traps. The most striking observational evidence for
dust trapping seems to come from large transitional disks, which feature giant
dust rings, sometimes lopsided, in which large quantities of dust appears to be
concentrated \citep{2013Natur.493..191C, 2013Sci...340.1199V}. These observations
appear to be well explained by the dust trapping scenario
\citet{2012A&A...545A..81P}. But these transitional disks seem to be rather
violent environments, possibly with strong warps \citep{2015ApJ...798L..44M,
  2017A&A...597A..42B} and companion-induced spirals
\citep{2016ApJ...816L..12D}. For more ``normal'' protoplanetary disk the dust
traps would have to be more subtle. \cite{2012A&A...538A.114P} explored the
possibility that the disk contains many axisymmetric local pressure maxima, and
calculated how the dust drift and growth would behave under such conditions. It
was found that, if the pressure bumps are strong enough, the dust trapping can
keep a sufficient fraction of the dust mass at large distances from the star to
explain the observed dust millimeter flux. It would leave, however, a detectable
pattern of rings that should be discernable with ALMA observations.  Since the
multi-ringed disk observation of HL Tau \citep{2015ApJ...808L...3A} a number of
such multi-ringed disks have been detected \citep{2016ApJ...820L..40A,
  2016PhRvL.117y1101I, 2017ApJ...851L..23C, 2017A&A...600A..72F,
  2018A&A...610A..24F}. It is therefore very tempting to see also these
multi-ringed disks as evidence for dust trapping, and as an explanation for
the retention of dust in the outer regions of protoplanetary disks.

@@

If we assume that the rings seen in our data are caused by dust trapping, the
question arises: what constraints do the data impose on the physics of dust
trapping? To explore this we start by studying the rings individually, assuming
that the dust does not escape from the ring. This makes it possible to look for
a steady-state dust trapping solution in which the radial drift forces (that push
the dust to the pressure peak) are balanced by turbulent mixing (that tends to
smear out the dust away from the pressure peak). In Section
\ref{sec-steady-state-analytic-trap-model} we will construct a very simplified
analytic dust trapping model, and confront this with the most well-isolated
rings from our sample. {\bf Describe the next sections.} But before we do the
modeling, we review, in Section \ref{sec-data}, the key aspects of the data from
our ALMA Large Programme that we compare to our models.

\section{The high-contrast rings of AS 209, Elias 24, HD 163296, GW Lup and HD 143006}
\label{sec-data}
%
In this paper we focus on those sources of the ALMA Large Programme
that show high-contrast and radially thin rings that are separated by deep
valleys, and that are sufficiently face-on to not have to worry about 3-D
line-of-sight issues. These are AS 209, Elias 24, HD 163296, GW Lup and
HD 143006. Their stellar parameters are given in Table \ref{tab-stellar-params}.

\begin{table}
\begin{center}
\begin{tabular}{|c|ccc|}
\hline
\hline
Source     & $d\,\mathrm{[pc]}$ & $M_{*}/M_{\odot}$ & $L_{*}/L_{\odot}$ \\
\hline
AS 209     & 121              & 0.912        &  0.631 \\
Elias 24   & 136              & 0.776        &  6.03  \\
HD 163296  & 101              & 2.14         &  25.7  \\
GW Lup     & 155              & 0.457        &  0.331 \\
HD 143006  & 166              & 0.813        &  0.871 \\
\hline
\hline
\end{tabular}
\end{center}
\caption{\label{tab-stellar-params}The stellar parameters for the
  stars studied in this paper. Distance is in parsec and mass and luminosity
  are in units of the solar values.}
\end{table}

A gallery of these sources is shown in Fig.~\ref{fig-obs-images}.
For an overview of the ALMA Large Programme we refer to {\bf Andrews et
  al.~XXXX}, and for an in-depth discussion on the data of the individual
sources we refer to {\bf [LIST HERE THE PAPERS]}.

\begin{figure*}
\centerline{\includegraphics[width=0.95\textwidth]{gallery_rings.pdf}}
\caption{\label{fig-obs-images}The continuum maps in
  band 6 of the five disks in our sample which have the most pronounced rings.
  The eight highest contrast rings, which are the topic of this paper,
  are marked in the images. For a detailed description of these
  data, see {\bf Guzm\'an et al.~(2018)} for AS 209, {\bf XXXX}}
\end{figure*}

The high-contrast rings of these sources provide ``clean laboratories''
for testing the theory of dust trapping, both in a ring-by-ring manner, as well
as by comparing to global dust drift models. Fig.\ \ref{fig-obs-profiles}
shows the radial profile (deprojected for
inclination) of the thermal emission of the dust of the five disks. The
procedure used to extract these radial profiles from the continuum maps
is described in {\bf [Paper Jane Huang, but note that for HD 163296 and
HD 143006 we removed the blob]}. The
brightness is expressed as linear brightness temperature
$T_{\mathrm{bright,linear}}=(c^2/2k_B)I_\nu/\nu^2$, where $I_\nu$ is the
intensity, $c$ the light speed, $k_B$ the Boltzmann constant, and $\nu$ the
frequency\footnote{It should be noted that the full (non-linear) brightness temperature,
defined such that $B_\nu(T_{\mathrm{bright,full}})=I_\nu$, with $B_\nu(T)$ the
Planck function, can have very different values, in particular at low
brightness. As a result $T_{\mathrm{bright,full}}(r)$ tends to vastly
underestimate the contrast between the rings and the inter-ring
regions. Therefore, in this paper we plot the brightness using the linear
brightness temperature, which is proportional to the intensity in Jy/beam.}.

\begin{figure*}
\centerline{\includegraphics[width=0.85\textwidth]{gallery_profiles.pdf}}
\caption{\label{fig-obs-profiles}The linear brightness temperature in
  band 6 of the five disks in our sample which have the most pronounced rings.
  The vertical axis is logarithmic to better show the contrast.
  The eight highest contrast rings are fitted by a Gaussian profile, shown
  as the solid inverse parabolas. The dotted inverse parabolas are Gaussians
  with the width of the ALMA beam. For a detailed description of these
  data, see {\bf Guzm\'an et al.~(2018)} for AS 209, {\bf XXXX}}
\end{figure*}

The eight rings which we study in this paper (see annotations in
Fig.~\ref{fig-obs-images}) are fitted with a Gaussian profile, shown as inverse
parabolas in Fig.\ \ref{fig-obs-profiles}. The fitting procedure was done in the
image plane, in order to make it possible to fit to a limited radial range,
i.e.\ selecting a single ring. Details of this fitting procedure, and tests
using the ALMA simulator, are described in Appendix {\bf XXXXX}. In
Fig.~\ref{fig-obs-gaussfits} these gaussian fits are shown in close-up. The
parameters of the best-fitting Gaussians are listed in Table
\ref{tab-gauss-params}.

\begin{figure*}
\centerline{\includegraphics[width=0.85\textwidth]{gaussfits_allrings.pdf}}
\caption{\label{fig-obs-gaussfits}Gaussian fits to the eight rings of this
  paper. The blue curves are the observations, the green curves are the best fit
  Gaussian profiles. The ``fit range'' bar shows the radial range within which
  the Gauss was fitted to the data. The fit range was chosen to fit the part of
  the curve that, by eye, most resembles a Gaussian. The ``beam'' bar shows the
  FWHM beam size of the observations. The grey band around the blue curve shows
  the estimated uncertainty of the data.}
\end{figure*}

\begin{figure*}
  \centerline{
    %\includegraphics[width=0.3\textwidth]{AS_209_doublegauss.pdf}
    \includegraphics[width=0.42\textwidth]{HD_163296_doublegauss.pdf}
    \includegraphics[width=0.42\textwidth]{HD_143006_doublegauss.pdf}
  }
  \caption{\label{fig-obs-doublegauss}The sum of the two Gaussian fits
    for the two sources with two partly overlapping rings: HD 163296 and
    HD 143006.}
\end{figure*}

\begin{table*}
\begin{center}
\begin{tabular}{|cc|cc|ccccc|ccccc|}
\hline
\hline
%Source & Ring & $r_i\, [\mathrm{au}]$ & $r_o\, [\mathrm{au}]$ & $A\, [\mathrm{K}]$ & $r\, [\mathrm{au}]$ & $\mathrm{fwhm}\, [\mathrm{mas}]$ & $\sigma\, [\mathrm{au}]$ & $w_d\, [\mathrm{au}]$ & $T_{\mathrm{model}}\,\mathrm{[K]}$ & $w_d/h_p$ & $M_{\mathrm{d,ri}} [M_{\oplus}]$ & $M_{\mathrm{d,iv}} [M_{\oplus}]$\\
Source & Ring & $r_{\mathrm{in}}$ & $r_{\mathrm{out}}$ & $A$ & $r_0$ & $\mathrm{fwhm}$ & $\sigma$ & $w_d$ & $T_{\mathrm{model}}$ & $T_{\mathrm{model,br,lin}}$ & $w_d/h_p$ & $M_{\mathrm{d,ri}}$ & $M_{\mathrm{d,iv}}$\\
            & & $[\mathrm{au}]$ & $[\mathrm{au}]$ & $[\mathrm{K}]$ & $[\mathrm{au}]$ & $[\mathrm{mas}]$ & $[\mathrm{au}]$ & $[\mathrm{au}]$ & $\mathrm{[K]}$ & $\mathrm{[K]}$ &  & $[M_{\oplus}]$ & $[M_{\oplus}]$\\
\hline
% NOTE: This table is made automatically with python_make_tables/make_latex_table.py
AS 209     & 1 &  69 &  79 &   3.64 &  74.13 &  75.42 &   3.88 &   3.43 &  12.9 &   8.2 &   0.7 &  44.4 &   2.41\\
AS 209     & 2 & 115 & 125 &   2.95 & 120.31 &  90.28 &   4.64 &   4.28 &  10.1 &   5.6 &   0.5 & 101.9 &   5.54\\
Elias 24   & 1 &  72 &  82 &   5.22 &  76.85 &  87.99 &   5.08 &   4.66 &  22.3 &  17.2 &   0.6 &  41.1 &   2.23\\
HD 163296  & 1 &  60 &  76 &   8.48 &  67.41 & 171.73 &   7.36 &   7.21 &  34.2 &  29.0 &   1.6 &  50.4 &   2.74\\
HD 163296  & 2 &  94 & 104 &   4.92 &  99.86 & 133.56 &   5.73 &   5.53 &  28.1 &  22.9 &   0.8 &  42.5 &   2.31\\
GW Lup     & 1 &  79 &  89 &   1.48 &  85.13 &  72.08 &   4.74 &   4.15 &  10.3 &   5.7 &   0.6 &  36.3 &   1.97\\
HD 143006  & 1 &  35 &  45 &   3.21 &  41.05 &  71.76 &   5.06 &   4.42 &  18.8 &  13.8 &   1.7 &  16.7 &   0.91\\
HD 143006  & 2 &  59 &  72 &   2.48 &  64.83 & 111.83 &   7.88 &   7.49 &  15.0 &  10.1 &   1.7 &  43.5 &   2.36\\
\hline
\hline
\end{tabular}
\end{center}
\caption{\label{tab-gauss-params}The model parameters for the Gaussian ring fits
  in Figs.~\ref{fig-obs-profiles} and \ref{fig-obs-gaussfits}. The values $r_{\mathrm{in}}$
  and $r_{\mathrm{out}}$ are the inner and outer boundary of the fitting domain. $A$ is the
  peak brightness temperature of the ring, $r_0$ is its radius in $\mathrm{au}$,
  $\mathrm{fwhm}$ is the width expressed as full-width-at-half-maximum in units
  of milliarcsecond, $\sigma$ is its width expressed as standard deviation in
  units of $\mathrm{au}$ and $w_d$ is the width of the underlying (deconvolved)
  dust emission profile, also expressed as standard deviation in units of
  $\mathrm{au}$. The last five columns are derived from a simple disk model
  assumption: $T_{\mathrm{model}}$ is the midplane temperature computed from
  Eq.~(\ref{eq-disk-temperature-model}) assuming a flaring angle of
  $\varphi=0.02$. From that one can derive the corresponding linear brightness
  temerature $T_{\mathrm{model,br,lin}}=B_\nu(T_{\mathrm{model}})c^2/2\nu^2$ with
  $\nu=c/0.13\,\mathrm{cm}$ is the frequency belonging to the wavelength of
  0.13 $\mathrm{cm}$ of ALMA band 6.
  The third-to-last column expresses the deconvolved dust ring
  width in units of the disk pressure scale height $h_p$ computed from
  $T_{\mathrm{model}}$. The last two columns are the computed dust mass in the
  ring in units of Earth masses, assuming the small grain absorption opacity
  $\kappa_{\mathrm{abs,ri}}(\lambda=0.13\,\mathrm{cm})=1.70\,\mathrm{cm}^2/\mathrm{g}$
  of \citet{2010A&A...512A..15R} ($M_{\mathrm{d,ri}}$) and the corresponding
  value
  $\kappa_{\mathrm{abs,ri}}(\lambda=0.13\,\mathrm{cm})=31.3\,\mathrm{cm}^2/\mathrm{g}$
  of \citet{1997MNRAS.291..121I} ($M_{\mathrm{d,iv}}$). The stellar parameter
  are taken from Table \ref{tab-stellar-params}.}
\end{table*}
